---
title       : 140.655 Biostat Lab 5
author      : L. Collado-Torres
framework   : minimal
highlighter : prettify
hitheme     : twitter-bootstrap
mode        : selfcontained
github      : {user: lcolladotor, repo: BiostatLab5, branch: gh-pages}
widgets     : [mathjax, disqus, ganalytics]
assets:
  css: 
    - "http://fonts.googleapis.com/css?family=PT+Sans"
    - "http://odyniec.net/articles/turning-lists-into-trees/css/tree.css"
---

```{r setup, echo = F, cache = F}
require("knitr")
opts_chunk$set(tidy=TRUE, width=50, echo=FALSE, tidy.opts=list(width.cutoff=50))
```





```{r "citations", echo=FALSE, message=FALSE}
## Load knitcitations with a clean bibliography
library("knitcitations")
cleanbib()
cite_options(tooltip=TRUE)

## I made my own citing function since citep() doesn't work like I want to with
## urls that are not really pages themselves like part of a GitHub repo.
mycitep <- function(x, short=NULL, year=substr(date(), 21, 24), tooltip=TRUE) {
	res <- tmp <- citep(x)
	if(!is.null(short)) {
		res <- gsub("></a>", paste0(">", short, "</a>"), tmp)
	}		
	if(tooltip) {
		res <- gsub("\\?\\?\\?\\?", year, res)
	}
	res <- gsub("span> ", "span>", res)
	res
}

## Here's an example
# mycitep("https://github.com/lcolladotor/lcollado753", "Collado, 2013")
```
```{r bibsetup, echo=FALSE, message=FALSE, warning=FALSE}
write.bibtex(c("knitcitations" = citation("knitcitations"), "slidify" = citation("slidify")), file = "pkgs.bib")
bib <- read.bibtex("pkgs.bib")
```

This project was done by [Emily Huang](http://www.jhsph.edu/departments/biostatistics/directory/students/phd.html) and [Leonardo Collado-Torres](http://bit.ly/LColladoTorres) for the [140.655 Analysis of Longitudinal Data](http://www.jhsph.edu/courses/course/140.655/01/2013/17988/) laboratory session 5 for biostatistics students.


## Instructions

```
Conduct, using a simulation, a sensitivity analysis of the change in effects
estimates as a function of the number of quadrature points.
Consider different integration methods.
```

As a reference/dataset starting point consider using `r mycitep("10.1111/1467-9876.00237")`.


## Theory

Below some of the key theoretical points are shown. For more details, please check out the [presentation](http://lcolladotor.github.io/BiostatLab5/GLMM.pdf).

### General GLMM setup

1) Distribution assumption
* $Y_{ij}|b_{i}\sim$ Exponential family
* $Var(Y_{ij}|b_{i})=v\{E(Y_{ij}|b_{i})\}\phi$, where $v$ is a known function
* $Cov(Y_{ij},Y_{ik}|b_{i})=0$

2) Systematic component
* $\eta_{ij}=X_{ij}\beta+Z_{ij}b_{i}$

3) Link function
* $g\{E(Y_{ij}|b_{i})\}=\eta_{ij}=X_{ij}\beta+Z_{ij}b_{i}$ for some known link function, $g$

4) Random effects
* Assumed to have some probability distribution, such as $b_{i}\sim MVN(0,G)$
* $b_{i}$ are assumed to be independent of the covariates


### Gauss-Hermite Quadrature

$$\int_{-\infty}^{\infty} & h(v)e^{-v^{2}}dv & \approx \sum_{k=1}^{d}h(x_{k})w_{k}$$

* $d$ quadrature points (weights, $w_{k}$, and evaluation points, $x_{k})$
* The more quadrature points used, the more accurate the approximation
* But computational burden increases with quadrature points, and grows exponentially with the number of random effects


## Simulation



## References

Web document generated using `slidify` `r mycitep(bib[["slidify"]])`. Citations made with `knitcitations` `r mycitep(bib[["knitcitations"]], "Boettiger, 2013")`. 


```{r bibliography, results='asis', echo=FALSE, cache=FALSE}
## Print bibliography
bibliography()
```



## R code

```{r "code", echo=TRUE, eval=FALSE}
```

## Reproducibility

This report was last updated on

```{r "date"}
Sys.time()
```

R session information:

```{r "reprod"}
sessionInfo()
```

```{r slidify, eval=FALSE, echo=TRUE}
## Generate this report
library("slidify")
slidify("index.Rmd")
```


<div id='disqus_thread'></div>


